\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-18 paper/main.tex
add %pip install davos to beginning of Fig 4, update line numbers throughout text

After importing \texttt{davos} (line 1), we first smuggle two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 3 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 4 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring that users have a particular exact version
installed would likely be unnecessarily restrictive. However, a
\textit{future} release might introduce some breaking change.  The
onion comment in line 4 helps ensure the analysis notebook continues
to run properly in the future, by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Line 6 then uses the \texttt{davos.config} object to enable
\texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy}~\cite{HarrEtal20} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Section~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Setting the \texttt{auto\_rerun} attribute to \texttt{True} also
benefits the (potential) installation of \texttt{pandas} on the next
lines:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
If we suppose that the data contained in \texttt{data-old.pkl} is
stored in a \texttt{Panel} object (i.e., created in a version of
\texttt{pandas} prior to 0.25.0), then we would not be able to load in
that object with newer versions of \texttt{pandas}.  Line 10 ensures
that an older version of \texttt{pandas} will be imported, enabling
the data to be read in (and, in line 12, saved out to a \texttt{.csv}
file, which is compatible with newer versions of \texttt{pandas}).

Newer versions of \texttt{pandas} have brought substantial
performance improvements, added new functionality, fixed bugs, etc.
Although the original dataset had to be read in using an older version
of the package, we can take advantage of those newer changes using a
new \texttt{smuggle} statement and onion comment on line 14 (which
specifies that version 1.3.5 should be imported).  Since
\texttt{pandas} had already been imported into the current workspace
(on line 10), the runtime must be restarted in order to gain access to
the newer version of \texttt{pandas}.  The \texttt{.auto\_rerun} flag
set on line 6 enables this process to happen automatically, and saving
out the dataset to a \texttt{.csv} file in lines 9--12 ensures that
the older version of \texttt{pandas} does not need to be reinstalled.

Next, line 16 uses the \texttt{davos.configure()} function to disable
the \texttt{auto\_rerun} option and simultaneously enable two other
options: \texttt{suppress\_stdout} and \texttt{noninteractive}. With
these options enabled, lines 17--18 \texttt{smuggle}
\texttt{TensorFlow}~\cite{AbadEtal15}, a powerful end-to-end platform
for building and working with machine learning models, and
\texttt{UMAP}~\cite{McInEtal18b}, a package that implements a family
of related manifold learning techniques. The onion comment in line 18
also specifies that \texttt{UMAP} should be installed with the
optional requirements needed for its ``plot'' and ``parametric\_umap''
features. Together, these two packages depend on 36 other unique
packages, most of which have dependencies of their own. And if many of
these are not already installed in the user's environment, lines
17--18 could take several minutes to run.  Enabling the
\texttt{noninteractive} option ensures that the installation will
continue automatically without user input during that time.  Enabling
\texttt{suppress\_stdout} also suppresses console outputs from the
installer that might otherwise distract from other more important outputs.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example4}
\end{center}

After re-enabling these two options (line 19), we next smuggle
specific versions of three plotting packages:
\texttt{Matplotlib}~\cite{Hunt07}, \texttt{seaborn}~\cite{Wask21}, and
\texttt{Quail}~\cite{HeusEtal17} (lines 21--23). Because the first two
are requirements of \texttt{UMAP}'s optional ``plot'' feature, they
will have already been installed by line 18, though possibly as
different versions than those specified in the onion comments on lines
21 and 22. If the installed and specified versions are the same, these
\texttt{smuggle} statements will function like standard \texttt{import}
statements to load the packages into the notebook namespace. If they
differ, \texttt{davos} will download the requested versions in place
of the installed versions before doing so.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example5}
\end{center}
Line 23 uses an onion comment to specify that \texttt{Quail} should be
installed directly from a specific GitHub commit (\texttt{6c847a4}).
This ability to install packages directly from GitHub repositories can
enable developers to use forked or modified versions of other
packages in their notebooks, even if those versions have not been
officially released.

In lines 25--28, we demonstrate another aspect of \texttt{davos}'s
functionality that supports more advanced installation scenarios.  The
\texttt{ipywidgets}~\cite{FredEtal15} package provides an API for
creating widgets, and the \texttt{widgetsnbextension} package provides
the machinery needed by the notebook frontend to display them.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example6}
\end{center}
A complication is that \texttt{ipywidgets} must be installed in the
same environment as the IPython kernel, whereas
\texttt{widgetsnbextension} must be installed in the environment that
houses the Jupyter notebook server. In standard setups, these two
environments are the same.  However, a common ``advanced'' approach
entails running the notebook server from a base environment, with
additional environments each providing their own separate,
interchangeable IPython kernels.  To accomodate this multi-environment
scenario, on lines 25 and 27 we control which environments each
package should be installed to.  Once these two packages are installed
and imported, line 30 smuggles \texttt{tqdm}~\cite{daCoEtal22}, which
display progress bars to provide status updates for running code. In
Jupyter notebooks, the \texttt{tqdm.notebook} module can be imported
to enable aesthetically pleasing progress bars that are displayed via
\texttt{ipywidgets}, if that package is installed and
importable. Therefore, to take advantage of this feature, it was
important to \texttt{smuggle} \texttt{tqdm} after ensuring the
\texttt{ipywidgets} package was available.

Next, we load in the reformatted dataset (line 32) and pre-trained
model (line 34) that we wish to use in our analysis.  In our
hypothetical example, we can suppose that the model was provided as a
\texttt{scikit-learn} \texttt{Pipeline} object that passes data
through two pretrained models in succession.  First, a trained \texttt{CountVectorizer}
instance converts text data to an array of word counts.  Second, the
word counts are passed to a topic model~\cite{BleiEtal03} using a
pretrained \texttt{LatentDirichletAllocation} instance.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
Let us suppose that the \texttt{Pipeline} object had been saved by its
original creator using the \texttt{joblib} package, as
\texttt{scikit-learn}'s documentation recommends.  Because
\texttt{joblib} uses the \texttt{pickle} protocol internally, the
ability to save and load pre-trained models is not guaranteed across
different \texttt{scikit-learn} versions.  For example, suppose that
the \texttt{Pipeline} object was created using \texttt{scikit-learn}
v0.21.3).  In that version of \texttt{scikit-learn}, the
\texttt{LatentDirichletAllocation} class is defined in
\texttt{sklearn.decomposition.\_lda}.  However, in version
0.22.0, that module was renamed to \texttt{\_online\_lda}, and in
versions 0.22.1 and higher, the name was reverted back to
\texttt{\_lda}.

In order to correctly load the model that includes the pretrained
\texttt{LatentDirichletAllocation} instance, in line 33, we first
smuggle a version of \texttt{scikit-learn} prior to v0.22.0 (i.e.,
before the first time the relevant module's name was changed).  Once
the model is loaded and reconstructed in memory from a compatible
package version (line 34), we upgrade to a newer version of
\texttt{scikit-learn} in line 35.  Taken together, the code in
Figure~\ref{fig:illustrative-example} shows how \texttt{davos} can
enable users to load in data and models that are incompatible with
newer versions of \texttt{pandas} and \texttt{scikit-learn}, but still
\textit{analyze} and manipulate the data and model output using the
latest approaches and implementations.

>>>>

After installing and importing \texttt{davos} (lines 1--2), we first smuggle two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 4 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 5 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring that users have a particular exact version
installed would likely be unnecessarily restrictive. However, a
\textit{future} release might introduce some breaking change.  The
onion comment in line 5 helps ensure the analysis notebook continues
to run properly in the future, by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
\texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy}~\cite{HarrEtal20} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Section~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Setting the \texttt{auto\_rerun} attribute to \texttt{True} also
benefits the (potential) installation of \texttt{pandas} on the next
lines:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
If we suppose that the data contained in \texttt{data-old.pkl} is
stored in a \texttt{Panel} object (i.e., created in a version of
\texttt{pandas} prior to 0.25.0), then we would not be able to load in
that object with newer versions of \texttt{pandas}.  Line 11 ensures
that an older version of \texttt{pandas} will be imported, enabling
the data to be read in (and, in line 13, saved out to a \texttt{.csv}
file, which is compatible with newer versions of \texttt{pandas}).

Newer versions of \texttt{pandas} have brought substantial
performance improvements, added new functionality, fixed bugs, etc.
Although the original dataset had to be read in using an older version
of the package, we can take advantage of those newer changes using a
new \texttt{smuggle} statement and onion comment on line 15 (which
specifies that version 1.3.5 should be imported).  Since
\texttt{pandas} had already been imported into the current workspace
(on line 11), the runtime must be restarted in order to gain access to
the newer version of \texttt{pandas}.  The \texttt{.auto\_rerun} flag
set on line 7 enables this process to happen automatically, and saving
out the dataset to a \texttt{.csv} file in lines 10--13 ensures that
the older version of \texttt{pandas} does not need to be reinstalled.

Next, line 17 uses the \texttt{davos.configure()} function to disable
the \texttt{auto\_rerun} option and simultaneously enable two other
options: \texttt{suppress\_stdout} and \texttt{noninteractive}. With
these options enabled, lines 18--19 \texttt{smuggle}
\texttt{TensorFlow}~\cite{AbadEtal15}, a powerful end-to-end platform
for building and working with machine learning models, and
\texttt{UMAP}~\cite{McInEtal18b}, a package that implements a family
of related manifold learning techniques. The onion comment in line 19
also specifies that \texttt{UMAP} should be installed with the
optional requirements needed for its ``plot'' and ``parametric\_umap''
features. Together, these two packages depend on 36 other unique
packages, most of which have dependencies of their own. And if many of
these are not already installed in the user's environment, lines
18--19 could take several minutes to run.  Enabling the
\texttt{noninteractive} option ensures that the installation will
continue automatically without user input during that time.  Enabling
\texttt{suppress\_stdout} also suppresses console outputs from the
installer that might otherwise distract from other more important outputs.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example4}
\end{center}

After re-enabling these two options (line 20), we next smuggle
specific versions of three plotting packages:
\texttt{Matplotlib}~\cite{Hunt07}, \texttt{seaborn}~\cite{Wask21}, and
\texttt{Quail}~\cite{HeusEtal17} (lines 22--24). Because the first two
are requirements of \texttt{UMAP}'s optional ``plot'' feature, they
will have already been installed by line 19, though possibly as
different versions than those specified in the onion comments on lines
22 and 23. If the installed and specified versions are the same, these
\texttt{smuggle} statements will function like standard \texttt{import}
statements to load the packages into the notebook namespace. If they
differ, \texttt{davos} will download the requested versions in place
of the installed versions before doing so.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example5}
\end{center}
Line 24 uses an onion comment to specify that \texttt{Quail} should be
installed directly from a specific GitHub commit (\texttt{6c847a4}).
This ability to install packages directly from GitHub repositories can
enable developers to use forked or modified versions of other
packages in their notebooks, even if those versions have not been
officially released.

In lines 26--29, we demonstrate another aspect of \texttt{davos}'s
functionality that supports more advanced installation scenarios.  The
\texttt{ipywidgets}~\cite{FredEtal15} package provides an API for
creating widgets, and the \texttt{widgetsnbextension} package provides
the machinery needed by the notebook frontend to display them.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example6}
\end{center}
A complication is that \texttt{ipywidgets} must be installed in the
same environment as the IPython kernel, whereas
\texttt{widgetsnbextension} must be installed in the environment that
houses the Jupyter notebook server. In standard setups, these two
environments are the same.  However, a common ``advanced'' approach
entails running the notebook server from a base environment, with
additional environments each providing their own separate,
interchangeable IPython kernels.  To accomodate this multi-environment
scenario, on lines 26 and 28 we control which environments each
package should be installed to.  Once these two packages are installed
and imported, line 30 smuggles \texttt{tqdm}~\cite{daCoEtal22}, which
display progress bars to provide status updates for running code. In
Jupyter notebooks, the \texttt{tqdm.notebook} module can be imported
to enable aesthetically pleasing progress bars that are displayed via
\texttt{ipywidgets}, if that package is installed and
importable. Therefore, to take advantage of this feature, it was
important to \texttt{smuggle} \texttt{tqdm} after ensuring the
\texttt{ipywidgets} package was available.

Next, we load in the reformatted dataset (line 33) and pre-trained
model (line 35) that we wish to use in our analysis.  In our
hypothetical example, we can suppose that the model was provided as a
\texttt{scikit-learn} \texttt{Pipeline} object that passes data
through two pretrained models in succession.  First, a trained \texttt{CountVectorizer}
instance converts text data to an array of word counts.  Second, the
word counts are passed to a topic model~\cite{BleiEtal03} using a
pretrained \texttt{LatentDirichletAllocation} instance.
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example7}
\end{center}
Let us suppose that the \texttt{Pipeline} object had been saved by its
original creator using the \texttt{joblib} package, as
\texttt{scikit-learn}'s documentation recommends.  Because
\texttt{joblib} uses the \texttt{pickle} protocol internally, the
ability to save and load pre-trained models is not guaranteed across
different \texttt{scikit-learn} versions.  For example, suppose that
the \texttt{Pipeline} object was created using \texttt{scikit-learn}
v0.21.3).  In that version of \texttt{scikit-learn}, the
\texttt{LatentDirichletAllocation} class is defined in
\texttt{sklearn.decomposition.online\_lda}.  However, in version
0.22.0, that module was renamed to \texttt{\_online\_lda}, and in
version 0.22.1, it was again renamed to \texttt{\_lda}.

In order to correctly load the model that includes the pre-trained
\texttt{LatentDirichletAllocation} instance, in line 34, we first
smuggle a version of \texttt{scikit-learn} prior to v0.22.0 (i.e.,
before the first time the relevant module's name was changed).  Once
the model is loaded and reconstructed in memory from a compatible
package version (line 35), we upgrade to a newer version of
\texttt{scikit-learn} in line 36.  Taken together, the code in
Figure~\ref{fig:illustrative-example} shows how \texttt{davos} can
enable users to load in data and models that are incompatible with
newer versions of \texttt{pandas} and \texttt{scikit-learn}, but still
\textit{analyze} and manipulate the data and model output using the
latest approaches and implementations.
